---
title: "Gapminder Multivariate Analysis"
author: "Ursula Neumann"
date: "March 2, 2018"
output: html_document
---


## Introduction
On the Gapminder website I found three datasets, one for female BMI over time ("Body Mass Index (BMI), women, Kg/m2"), one for total cholesterol levels over time ("Cholesterol (fat) in blood, women, mmol/L"), and one for systolic blood pressure over time (Blood pressure (SBP), women, mmHg).  These datasets are provided by the MRC-HPA Center for Environment and Health and span the years 1980 - 2008 for 199 countries.  I previously performed an analysis of cholesterol vs BMI, but now want to add in the third parameter blood pressure.  In addition, I would like to separate the data based on continent to see if any interesting observations can be made.

## Load datasets and prepare the dataframe for analysis
Below I have loaded all the datasets and transformed them into "tall" datasets using the gather function.  Then I performed 3 sequential inner joins to produce a dataframe with the columns: country, continent, year, BMI, cholesterol and SBP.  I'm unsure if there's a way to streamline this process but it was simple enough.


```{r}
# Load the necessary libraries
library("tidyr")
library("dplyr")
library("ggplot2")
library("gridExtra")

# Load the BMI dataset 
df1 <- read.csv('female_BMI.csv', check.names = FALSE)
head(df1)

# Gather the data into columns of country, year, and BMI
df_BMI <- gather(df1, key = "year", value = "BMI", 2:dim(df1)[2])
head(df_BMI)

# Load the cholesterol dataset
df2 <- read.csv('female_cholesterol.csv', check.names = FALSE)
head(df2)

# Gather the data into columns of country, year, and cholesterol
df_chol <- gather(df2, key = "year", value = "cholesterol", 2:dim(df2)[2])
colnames(df_chol)[1] <- "Country"
head(df_chol)

# Load the blood pressure dataset
df3 <- read.csv('female_blood_pressure.csv', check.names = FALSE)
head(df3)

# Gather the data into columns of country, year, and blood pressure
df_SBP <- gather(df3, key = "year", value = "SBP", 2:dim(df2)[2])
colnames(df_SBP)[1] <- "Country"
head(df_SBP)

# Load the countries-continents dataset
df4 <- read.csv('countries_continents.csv', check.names = FALSE)
head(df4)

# Join the BMI and cholesterol datasets using an inner join on country and year
df5 <- inner_join(df_BMI, df_chol, by = c("Country", "year"))
head(df5)

# Join the blood pressure dataset using an inner join on country and year
df6 <- inner_join(df5, df_SBP, by = c("Country", "year"))
head(df6)

# Join in the countires_continents dataset
df <- inner_join(df6, df4, by = c("Country"))
head(df)

# Reorder the columns
df <- df[c(1,6, 2, 3, 4, 5)]
head(df)

# Change the year variable to a numeric class
df$year <- as.numeric(df$year)
```

## Previous analysis: Cholesterol vs BMI
As shown below there is a moderately strong positive correlation (r = 0.51) between cholesterol and BMI.  


```{r}
# Graph a scatterplot of cholesterol vs BMI in 2008 for all countries 
ggplot(aes(x = BMI, 
           y = cholesterol), 
       data = subset(df, df$year == 2008)) + 
  geom_point(alpha = 1/10) +
  geom_smooth(method = 'lm', color = 'magenta') +
  labs(x = expression(paste("BMI (kg/m"^2, ")")),
       y = expression(paste("Total Cholesterol (mmol/L)")))
```


## Cholesterol vs BMI, colored by continent, point size by SBP
To add more information/variables to the graph above, I decided to color the points by continent, and size the points by blood pressure.  Below is a plot with all countries on one graph, as well as a plot where the points faceted by continent to be able to better view each cluster.  Africa and Asia have the most widespread BMI and cholesterol levels.  The BMIs of Europe, North America, and South America are more clustered in the middle of the graph.  I expected that the countries with the highest blood pressure would be those at the top right corner of the graph (high cholesterol and high BMI).  However, the largest points are those clustered at the bottom left corner of the graph, from African countries.

```{r}
ggplot(aes(x = BMI, 
           y = cholesterol), 
       data = subset(df, df$year == 2008)) + 
  geom_point(alpha = 1/3, aes(color = Continent, size = SBP)) +
  labs(x = expression(paste("BMI (kg/m"^2, ")")),
       y = expression(paste("Total Cholesterol (mmol/L)")))
```

```{r}
ggplot(aes(x = BMI, 
           y = cholesterol), 
       data = subset(df, df$year == 2008)) + 
  geom_point(alpha = 1/3, aes(color = Continent, size = SBP)) +
  labs(x = expression(paste("BMI (kg/m"^2, ")")),
       y = expression(paste("Total Cholesterol (mmol/L)")))+
  facet_wrap(~Continent)
```

## Resources
1) https://www.gapminder.org/data/
2) https://old.datahub.io/dataset/countries-continents/resource/aa08c34c-57e8-4e15-bd36-969bee26aba5
